TTmanuscript

### 2020-04-18 text/manuscript.Rmd
add references.bib; modify main text with bibtex refs

---
title: "TTmanuscript"
author: "Dennis Hazelett"
date: "4/18/2020"
output: html_document
---

@@

### Authors: Peter T. Nguyen, Simon G. Coetzee, Corona I. Rosario, and Dennis J. Hazelett

@@

Some considerable progress has been made in analyzing these data. Statistical analyses identify hundreds of global and tissue-specific cancer driver genes (pmid 22759861, 23884480, 24390350, 26209800, 27911828, 31287869, 31363082) using approaches aimed at detecting when genes are mutated at a greater rate than expected due to chance. Based on these studies, it has been proposed that on the order of three mutations in key oncogenes and/or tumor suppressors would be sufficient to transform a normal cell to a cancerous state (pmid 26559569).

Other studies have been aimed at providing a comprehensive view of mutations, gene expression and genomic signatures, with the express goal of understanding the common themes of all cancer independent of tissue of origin. The first of these (the “Pan-Cancer-12” analysis) identified a clade of 11 subtypes from 12 cancer types, using integrative analysis with co-equal weighting of gene expression, methylation, copy number and proteomics data (pmid 25109877). The principal finding was that tissue-of-origin is the predominant driving factor, though a significant proportion of tumors (10%) could be reclassified independent of tissue-of-origin. In a second study involving 33 cancer types and a much greater number of tumors using a newer iCluster method (pmid 19759197), the same authors identify 28 clusters that could be further related or subdivided into organ specific groups, including pan-gastrointestinal, pan-gynecological, pan-squamous, pan-gynecological/squamous and pan-kidney (pmid 29625048). The authors of this study analyzed these categories for pathway enrichment and noted multiple overlapping similarities between these latter categories.

More recently, attempts have been made to consider pan-cancer mutations and other data outside the context of enrichment at the gene level. Horn et al (pmid 29200198) introduce approaches to address the sparseness of mutations, allowing the potential importance of a gene to be influenced by the aggregate of positive selection in its nearest network neighbors. This method proves powerful especially when a small number (on the order of hundreds) of tumors are involved. However, the larger problem of cancer etiology remains: what mutations are required, in what order, and are there specific combinations of delinquent cellular processes sufficient to cause cancer? If there is a common logic to cancer, how is it to be detected underneath the overwhelming cell-of-origin signal?

@@

The basic problem in tumor research of this kind, namely that the search space, ~20,000 protein-coding genes is greater than the number of tumor samples (pmid 29200198). And that is not accounting for the fact that >30k non-coding genes and hundreds of thousands of non-coding regulatory elements exist with oncogenic or tumor suppressor potential. The bioinformatic challenge therefore is to define the problem precisely and then create a new dataset that addresses this problem. Hence, we define cancer as a malady of biochemical processes and pathways, and generated a dataset that scores each of these essential pathways as mutated or not.

It is important that we chose point mutations in exome sequencing data for this task because we know unambiguously in most cases the target gene. In addition, we reduce pathway state to a boolean value because it does not matter for our purposes whether a single tumor is “enriched” for pathway mutations, chiefly because such metrics favor larger pathways over smaller ones. The data are likely to be noisy for several reasons. First, point mutations can be deleterious (attenuating or hypomorphic) or activating (neomorphic) in genes that are potentially either activators or repressors. In the current model, we accept this uncertainty and assume mutations are generically disruptive to the pathway. Second, low-expressed genes and non-expressed genes are known to accumulate mutations at a higher rate (pmid 24132290, 29662161). To address this issue, we identified genes with low expression average for each type of cancer and eliminated them for that cancer only. Highly expressed genes also have high mutation rates owing to transcription induced mutagenesis (pmid 23146897). We felt that this mechanism results in cell-type-specific biases that might be biologically meaningful for predisposition to different classes of cancer however, and therefore chose to include these genes in our analysis.

@@

>>>>

---
title: "TTmanuscript"
author: "Dennis Hazelett"
date: "4/18/2020"
output: html_document
bibliography: references.bib
---

@@

#### Authors: Peter T. Nguyen, Simon G. Coetzee, Daniel Lakeland and Dennis J. Hazelett

@@

Some considerable progress has been made in analyzing these data. Statistical analyses identify hundreds of global and tissue-specific cancer driver genes [@deesMuSiC2012; @tamboreroOncodriveCLUST2013; @lawrenceDiscoverySaturationAnalysis2014; @kumarStatisticallyIdentifyingTumor2015; @tokheimEvaluatingEvaluationCancer2016; @jiangWITER2019; @zhaoModelingPositive2019] using approaches aimed at detecting when genes are mutated at a greater rate than expected due to chance. Based on these studies, it has been proposed that on the order of three mutations in key oncogenes and/or tumor suppressors would be sufficient to transform a normal cell to a cancerous state [@vogelsteinPathCancerThree2015].

Other studies have been aimed at providing a comprehensive view of mutations, gene expression and genomic signatures, with the express goal of understanding the common themes of all cancer independent of tissue of origin. The first of these (the “Pan-Cancer-12” analysis) identified a clade of 11 subtypes from 12 cancer types, using integrative analysis with co-equal weighting of gene expression, methylation, copy number and proteomics data [@hoadleyMultiplatformAnalysis2014]. The principal finding was that tissue-of-origin is the predominant driving factor, though a significant proportion of tumors (10%) could be reclassified independent of tissue-of-origin. In a second study involving 33 cancer types and a much greater number of tumors using a newer iCluster method [@shenIntegrativeClusteringMultiple2009], the same authors identify 28 clusters that could be further related or subdivided into organ specific groups, including pan-gastrointestinal, pan-gynecological, pan-squamous, pan-gynecological/squamous and pan-kidney [@hoadleyCellofOriginPatternsDominate2018]. The authors of this study analyzed these categories for pathway enrichment and noted multiple overlapping similarities between these latter categories.

More recently, attempts have been made to consider pan-cancer mutations and other data outside the context of enrichment at the gene level. Horn et al [@hornNetSigNetworkbasedDiscovery2018] introduce approaches to address the sparseness of mutations, allowing the potential importance of a gene to be influenced by the aggregate of positive selection in its nearest network neighbors. This method proves powerful especially when a small number (on the order of hundreds) of tumors are involved. However, the larger problem of cancer etiology remains: what mutations are required, in what order, and are there specific combinations of delinquent cellular processes sufficient to cause cancer? If there is a common logic to cancer, how is it to be detected underneath the overwhelming cell-of-origin signal?

@@

The basic problem in tumor research of this kind, namely that the search space, ~20,000 protein-coding genes is greater than the number of tumor samples [@hornNetSigNetworkbasedDiscovery2018]. And that is not accounting for the fact that >30k non-coding genes and hundreds of thousands of non-coding regulatory elements exist with oncogenic or tumor suppressor potential. The bioinformatic challenge therefore is to define the problem precisely and then create a new dataset that addresses this problem. Hence, we define cancer as a malady of biochemical processes and pathways, and generated a dataset that scores each of these essential pathways as mutated or not.

It is important that we chose point mutations in exome sequencing data for this task because we know unambiguously in most cases the target gene. In addition, we reduce pathway state to a boolean value because it does not matter for our purposes whether a single tumor is “enriched” for pathway mutations, chiefly because such metrics favor larger pathways over smaller ones. The data are likely to be noisy for several reasons. First, point mutations can be deleterious (attenuating or hypomorphic) or activating (neomorphic) in genes that are potentially either activators or repressors. In the current model, we accept this uncertainty and assume mutations are generically disruptive to the pathway. Second, low-expressed genes and non-expressed genes are known to accumulate mutations at a higher rate [@kandothMutationalLandscapeSignificance2013; @kimPancancerAnalysisSomatic2018]. To address this issue, we identified genes with low expression average for each type of cancer and eliminated them for that cancer only. Highly expressed genes also have high mutation rates owing to transcription induced mutagenesis [@parkGenomicEvidenceElevated2012]. We felt that this mechanism results in cell-type-specific biases that might be biologically meaningful for predisposition to different classes of cancer however, and therefore chose to include these genes in our analysis.

@@

## References:
